---
title: "Poisson_Binomial_Approx"
output: html_document
date: "2026-06-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploration: The Poisson Approximation of the Binomial Distribution
### Problem Context
The Binomial distribution can be approximated by a Poisson distribution when the number of trials $n$ becomes large and the probability of success $p$ becomes small, keeping the rate $\lambda = np$ constant.
We know the probability mass function (PMF) of the Binomial distribution is:$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$
Taking the limit as $n$ approaches infinity and $p$ approaches zero yields the Poisson PMF:$$\lim_{n \to \infty, p \to 0} \binom{n}{k} p^k (1-p)^{n-k} = \frac{\lambda^k e^{-\lambda}}{k!}$$ 
However, I became suspicious when the textbook stated rigid, specific bounds for which this is considered a "good" approximation (as stated in the source, $\lambda < 7$). This project empirically tests those claims to see where the approximation actually holds up.

### The Objective
We will simulate and compare Binomial distributions against their Poisson counterparts using varying parameters. Specifically, we will write a loop to calculate the Total Variation Distance ($d_{TV}$) for each setup, split the plotting screen to visually compare $\lambda = 4$, $\lambda = 6$, and $\lambda = 12$, display their corresponding distances below the graphs, and draw conclusions based on the empirical error data.


```{r, fig.height=4, fig.width=12}
# Set up variables  
n <- 100
lambdas <- c(4, 6, 12)
p_values <- lambdas / n
k_max <- 25 

# Set up transparent pastel colours
color_binom <- rgb(0.30, 0.71, 0.67, alpha = 0.6) 
color_pois <- rgb(1.00, 0.54, 0.40, alpha = 0.6)  

# Split the plotting screen
par(mfrow = c(1, 3))

# Create a for loop for each lambda
for (i in 1:3) {
  lambda <- lambdas[i]
  p <- p_values[i]
  k <- 0:k_max
  
  # 1. Calculate the p.m.f to find the TVD
  binom_probs <- dbinom(k, size = n, prob = p)
  pois_probs <- dpois(k, lambda = lambda)
  
  tvd <- 0.5 * sum(abs(binom_probs - pois_probs))
  y_max <- max(c(binom_probs, pois_probs))
  
  # 2. Plot Base for the Binomial
  barplot(binom_probs, 
          names.arg = k, 
          col = color_binom, 
          main = paste("Lambda =", lambda),
          sub = paste("TVD =", round(tvd, 4)),
          xlab = "k", 
          ylab = "Probability",
          ylim = c(0, y_max * 1.1),
          space = 0)
  
  # 3. Plot Overlay for the Poisson
  barplot(pois_probs, 
          col = color_pois, 
          border = NA,
          add = TRUE,
          space = 0)
  
  # Add the legend
  legend("topright",
         legend = c("Binomial", "Poisson"),
         fill = c(color_binom, color_pois),
         bty = "n")
}


```

### Conclusion: Evaluating the Approximation
To quantify how well the approximation actually holds, this project used the Total Variation Distance ($d_{TV}$):

$$d_{TV} = \frac{1}{2} \sum_{k=0}^{\infty} |P(X_{Bin}=k) - P(X_{Pois}=k)|$$

This measures the largest possible difference in probability the two distributions assign to any event, $0$ being a perfect match, $1$ being entirely disjoint. The natural follow-up question is: what counts as *good enough*? A quick Google search suggested that a $d_{TV} < 0.05$ is a reasonable threshold for practical use. Under that standard, pushing $\lambda$ all the way to $12$ ($p = 0.12$) — well outside the textbook's recommended bounds has still yielded a $d_{TV} \approx 0.031$. The maximum error on any single probability stays within a $3\%$ tolerance, which suggests the $\lambda < 7$ rule might be more conservative than strictly necessary.

### Further Comment 
One thing I noticed across all three plots: the Binomial consistently overshoots the Poisson around the mean, then undershoots it further out in the tails. This turns out to be a direct algebraic consequence of the variance gap:

$$\text{Var}(X_{Bin}) = \lambda(1-p) < \lambda = \text{Var}(X_{Pois})$$

Since $p > 0$, the Binomial always has a strictly tighter spread than its Poisson counterpart, so the same probability mass gets squeezed inward, producing a taller peak at the centre at the cost of lighter tails.

